An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-06-09 cal_demo/ms.tex
Switched around figure filenames after switching sections
Added text re:noise of data solutions, updated freq averaging in data

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig9.pdf}
\caption{Gain estimate errors as a function of integration time, receiver noise (line color), and 
calibration method. The solid lines represent the error on the EPICal derived gain estimates, 
while the dashed lines represent the error on the visibility based estimates. The thicker lines 
were derived using a full sky model in the calibration loops, and the thinner lines used only a 
single point source model. In the full sky model case we see all methods trend down with 
longer integration time as expected. For a given integration time, EPICal derived errors are on 
average 23\% higher than those of visibility based calibration. In the case of a single source 
model, the errors bottom out due to the flux in the measurements that is not modeled. In this 
regime the EPICal method performs slightly better than the visibility-based method due to 
effectively beamforming to the sky pixel where the incomplete model is most accurate. On 
average EPICal errors are 5\% lower in this regime.
}
\label{fig:errors}
\end{center}
\end{figure}

@@

We made several choices while studying the behavior of the LWA data to improve our 
calibration. Through our previous imaging work, we noted that the flux scale of uncalibrated 
images was consistent with average gain amplitudes of 0.25. To allow the calibration to 
converge quickly we initialized our gain estimates at this level. 
We also found a boost in signal to noise is achieved easily by assuming the gains are
constant across a range of frequencies, and averaging solutions across channels.
Here we average solutions across 150 channels, or about 29 kHz. 
With a fractional bandwidth $B/f_0 = 3.9 \times 10^{-4} \ll 1$, we assume a smooth bandpass 
across the band. 
A damping factor of $\gamma = 0.7$ was adopted.
The large damping factor, relative to what was used in simulations, was necessary due to
the larger noise levels in the LWA data compared to the ideal simulations.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig6.pdf}
\caption{Gain phase solutions as a function of calibration iteration for an LWA TBN observation. 
The gain estimates are initialized with zero phase, but quickly span a $2\pi$ range, and settle 
into relatively flat, albeit noisy, solutions. The majority of phases congregate near zero, which is 
not surprising given the fairly good quality image produced from uncalibrated data.
}
\label{fig:data_phase}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig7.pdf}
\caption{Gain amplitudes as a function of calibration iteration for an LWA TBN observation. The 
gains estimates were initialized with amplitude 0.25 after inspection of the electric field values 
compared to our model sky. The solutions are noisy, but flat. The range in amplitudes is due to 
the non-uniformity of cables between the LWA antennas and receivers.
}
\label{fig:data_amp}
\end{center}
\end{figure}

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The 
left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent 
near the center of the image, and Cas~A is also clearly visible in the upper right. The middle 
panel shows the image produced after calibration with identical integration time and bandwidth. 
The sidelobes throughout the image are significantly suppressed, and the galactic plane is 
much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature 
just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux 
from the global sky model (GSM, \citealt{deo08}). 
For reference we should the GSM \citep{pri16} in the right panel of 
Figure~\ref{fig:data_images}, convolved by the LWA point spread function and weighted
by  two factors of the primary beam to match the data images.

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{fig8a.pdf}
\includegraphics[width=0.3\linewidth]{fig8b.pdf}
\includegraphics[width=0.3\linewidth]{fig8c.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images 
were produced with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows 
significant reduction in rumble throughout, while retaining the prominent Cyg~A and 
Cas~A sources. The galactic plane is also substantially more evident after calibration. For 
reference, the GSM is shown in the 
right panel convolved with the LWA point spread function and weighted by two factors
of the primary beam to place it in the same holographic frame as the data images. While much of the GSM is visible in the calibrated image, we 
note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

@@

\textbf{Fitting gain models.} With some knowledge of the instrumental bandpass, the noise of 
the gain solutions can be greatly reduced by fitting a model to the per-frequency solutions 
derived here. We used this method at a rudimentary level in our demonstration to LWA data by 
assuming the gains were constant over a narrow bandwidth. One could easily improve on this 
by extending the bandwidth and fitting for a low order polynomial in phase and amplitude. 
Additionally, with knowledge of the poly-phase filters applied to the data stream, we could 
include channels closer to the edge of the band.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig6.pdf}
\caption{Gain estimate errors as a function of integration time, receiver noise (line color), and 
calibration method. The solid lines represent the error on the EPICal derived gain estimates, 
while the dashed lines represent the error on the visibility based estimates. The thicker lines 
were derived using a full sky model in the calibration loops, and the thinner lines used only a 
single point source model. In the full sky model case we see all methods trend down with 
longer integration time as expected. For a given integration time, EPICal derived errors are on 
average 23\% higher than those of visibility based calibration. In the case of a single source 
model, the errors bottom out due to the flux in the measurements that is not modeled. In this 
regime the EPICal method performs slightly better than the visibility-based method due to 
effectively beamforming to the sky pixel where the incomplete model is most accurate. On 
average EPICal errors are 5\% lower in this regime.
}
\label{fig:errors}
\end{center}
\end{figure}

@@

We made several choices while studying the behavior of the LWA data to improve our 
calibration. Through our previous imaging work, we noted that the flux scale of uncalibrated 
images was consistent with average gain amplitudes of 0.25. To allow the calibration to 
converge quickly we initialized our gain estimates at this level. 
We also found a boost in signal to noise is achieved easily by assuming the gains are
constant across a range of frequencies, and averaging solutions across channels.
Here we average solutions across 300 channels, or about 58.6 kHz. 
With a fractional bandwidth $B/f_0 = 7.9 \times 10^{-4} \ll 1$, we assume a smooth bandpass 
across the band. 
A damping factor of $\gamma = 0.7$ was adopted.
The large damping factor, relative to what was used in simulations, was necessary due to
the larger noise levels in the LWA data compared to the ideal simulations.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig7.pdf}
\caption{Gain phase solutions as a function of calibration iteration for an LWA TBN observation. 
The gain estimates are initialized with zero phase, but quickly span a $2\pi$ range, and settle 
into relatively flat, albeit noisy, solutions. The majority of phases congregate near zero, which is 
not surprising given the fairly good quality image produced from uncalibrated data.
}
\label{fig:data_phase}
\end{center}
\end{figure}

@@

To ensure the level of noise in our solutions is reasonable we compared to the fractional errors
from our simulations shown in Figure~\ref{fig:errors}. We first use the Global Sky Model
\citep{deo08} to generate a mock observed sky at 74.03 MHz using the PyGSM python 
module\footnote{https://github.com/telegraphic/PyGSM} \citep{pri16}. We integrate this model
to estimate a total sky flux power of 171,000~Jy, meaning our calibration pixel containing
Cyg A accounts for  approximately $20,539/171,000 \approx 12\%$ of the total sky power. 
This places us in a regime very similar to the incomplete sky model simulations of the 
previous section, where the calibration pixel accounted for 21.7\% of the sky power. We 
should therefore expect our fractional gain noise levels to be at best around 0.1, the level 
to which the simulations asymptotically approached.

To estimate our fractional gain noise we calculate the standard deviation of the gain
estimates after iteration 15 for each antenna separately. We then divide these standard
deviations by the respective mean gain amplitudes. This resulted in a distribution of fractional
gain noise centered at 0.15 and a width of 0.04 ($\approx 0.15/\sqrt{15}$). This noise level
is consistent with what we expected from the simulations.
There were four outliers with noise levels significantly outside the distribution 
($\sigma_g \approx 0.29$). Upon closer inspection the gain amplitudes of these outliers had
converged later in the calibration loop, after iteration 15.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig8.pdf}
\caption{Gain amplitudes as a function of calibration iteration for an LWA TBN observation. The 
gains estimates were initialized with amplitude 0.25 after inspection of the electric field values 
compared to our model sky. The solutions are noisy, but flat. The range in amplitudes is due to 
the non-uniformity of cables between the LWA antennas and receivers.
}
\label{fig:data_amp}
\end{center}
\end{figure}

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The 
left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent 
near the center of the image, and Cas~A is also clearly visible in the upper right. The middle 
panel shows the image produced after calibration with identical integration time and bandwidth. 
The sidelobes throughout the image are significantly suppressed, and the galactic plane is 
much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature 
just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux 
from the GSM. 
For reference we show the GSM in the right panel of 
Figure~\ref{fig:data_images}, convolved by the LWA point spread function and weighted
by  two factors of the primary beam to match the holographic frame of the data images.

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{fig9a.pdf}
\includegraphics[width=0.3\linewidth]{fig9b.pdf}
\includegraphics[width=0.3\linewidth]{fig9c.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images 
were produced with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows 
significant reduction in rumble throughout, while retaining the prominent Cyg~A and 
Cas~A sources. The galactic plane is also substantially more evident after calibration. For 
reference, the GSM is shown in the 
right panel convolved with the LWA point spread function and weighted by two factors
of the primary beam to place it in the same holographic frame as the data images. While much of the GSM is visible in the calibrated image, we 
note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

@@

\textbf{Fitting gain models.} With some knowledge of the instrumental bandpass, the noise of 
the gain solutions can be greatly reduced by fitting a model to the per-frequency solutions 
derived here. We used this method at a rudimentary level in our demonstration to LWA data by 
assuming the gains were constant over a narrow bandwidth. One could easily improve on this 
by extending the bandwidth and fitting for a low order polynomial in phase and amplitude. 
Additionally, with knowledge of 
any filters applied to the data stream in the receiver chain,
we could include channels closer to the edge of the band.
